---
title: "Practical Machine Learning - Course Project"
output: html_document
---

Introduction
For this project, we are given data from accelerometers on the belt, forearm, arm, and dumbell of 6 research study participants. Our training data consists of accelerometer data and a label identifying the quality of the activity the participant was doing. Our testing data consists of accelerometer data without the identifying label. Our goal is to predict the labels for the test set observations.

Below is the code I used when creating the model, estimating the out-of-sample error, and making predictions. I also include a description of each step of the process.

setting working directory

```{r}
setwd("C:/Users/tarun/Desktop/r/Practical_machine_learning")
```
Data Preparation
I load the caret package, and read in the training and testing data:

```{r}
library(AppliedPredictiveModeling)
library(caret)
library(rpart.plot)
library(rattle)
library(randomForest)
library(rpart)
library(RColorBrewer)

```

Getting the data
The training data set can be found on the following URL:

```{r}
trainUrl <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
```

The testing data set can be found on the following URL:
```{r}
testUrl <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
```

Load data to memory solely

```{r}
training <- read.csv(url(trainUrl), na.strings=c("NA","#DIV/0!",""))
testing <- read.csv(url(testUrl), na.strings=c("NA","#DIV/0!",""))
```

Partioning the training set into two
Partioning Training data set into two data sets, 60% for myTraining, 40% for myTesting:

```{r}
inTrain <- createDataPartition(y=training$classe, p=0.6, list=FALSE)
myTraining <- training[inTrain, ]
myTesting <- training[-inTrain, ]
dim(myTraining)
dim(myTesting)
```

Cleaning the data
The following transformations were used to clean the data:

Transformation 1: Cleaning NearZeroVariance Variables Run this code to view possible NZV Variables:
```{r}
myNZVvars <- names(myTraining) %in% c("new_window", "kurtosis_roll_belt", "kurtosis_picth_belt",
"kurtosis_yaw_belt", "skewness_roll_belt", "skewness_roll_belt.1", "skewness_yaw_belt",
"max_yaw_belt", "min_yaw_belt", "amplitude_yaw_belt", "avg_roll_arm", "stddev_roll_arm",
"var_roll_arm", "avg_pitch_arm", "stddev_pitch_arm", "var_pitch_arm", "avg_yaw_arm",
"stddev_yaw_arm", "var_yaw_arm", "kurtosis_roll_arm", "kurtosis_picth_arm",
"kurtosis_yaw_arm", "skewness_roll_arm", "skewness_pitch_arm", "skewness_yaw_arm",
"max_roll_arm", "min_roll_arm", "min_pitch_arm", "amplitude_roll_arm", "amplitude_pitch_arm",
"kurtosis_roll_dumbbell", "kurtosis_picth_dumbbell", "kurtosis_yaw_dumbbell", "skewness_roll_dumbbell",
"skewness_pitch_dumbbell", "skewness_yaw_dumbbell", "max_yaw_dumbbell", "min_yaw_dumbbell",
"amplitude_yaw_dumbbell", "kurtosis_roll_forearm", "kurtosis_picth_forearm", "kurtosis_yaw_forearm",
"skewness_roll_forearm", "skewness_pitch_forearm", "skewness_yaw_forearm", "max_roll_forearm",
"max_yaw_forearm", "min_roll_forearm", "min_yaw_forearm", "amplitude_roll_forearm",
"amplitude_yaw_forearm", "avg_roll_forearm", "stddev_roll_forearm", "var_roll_forearm",
"avg_pitch_forearm", "stddev_pitch_forearm", "var_pitch_forearm", "avg_yaw_forearm",
"stddev_yaw_forearm", "var_yaw_forearm")
myTraining <- myTraining[!myNZVvars]
dim(myTraining)
```

Transformation 2: Killing first column of Dataset - ID Removing first ID variable so that it does not interfer with ML Algorithms:

```{r}
myTraining <- myTraining[c(-1)]
```

Transformation 3: Cleaning Variables with too many NAs. For Variables that have more than a 60% threshold of NA’s I’m going to leave them out:


```{r}
trainingV3 <- myTraining #creating another subset to iterate in loop
for(i in 1:length(myTraining)) { 
  #for every column in the training dataset
        if( sum( is.na( myTraining[, i] ) ) /nrow(myTraining) >= .6 ) { #if n?? NAs > 60% of total observations
        for(j in 1:length(trainingV3)) {
            if( length( grep(names(myTraining[i]), names(trainingV3)[j]) ) ==1)  { #if the columns are the same:
                trainingV3 <- trainingV3[ , -j] #Remove that column
            }   
        } 
        }
}
#To check the new N?? of observations
  dim(trainingV3)
    ```
#Seting back to our set:

```{r}
  myTraining <- trainingV3
rm(trainingV3)
```
Now let us do the exact same 3 transformations but for our myTesting and testing data sets.

```{r}
clean1 <- colnames(myTraining)
clean2 <- colnames(myTraining[, -58]) #already with classe column removed
myTesting <- myTesting[clean1]
testing <- testing[clean2]
```
#To check the new N?? of observations
```{r}
dim(myTesting)
```
#To check the new N?? of observations
```{r}
dim(testing)
```
#Note: The last column - problem_id - which is not equal to training sets, was also "automagically" removed
#No need for this code:
#testing <- testing[-length(testing)]



In order to ensure proper functioning of Decision Trees and especially RandomForest Algorithm with the Test data set (data set provided), we need to coerce the data into the same type.



```{r}
for (i in 1:length(testing) ) {
        for(j in 1:length(myTraining)) {
        if( length( grep(names(myTraining[i]), names(testing)[j]) ) ==1)  {
            class(testing[j]) <- class(myTraining[i])
        }      
    }      
}
```

#And to make sure Coertion really worked, simple smart ass technique:
```{r}
testing <- rbind(myTraining[2, -58] , testing) #note row 2 does not mean anything, this will be removed right.. now:
testing <- testing[-1,]
```


Using ML algorithms for prediction: Decision Tree
```{r}
modFitA1 <- rpart(classe ~ ., data=myTraining, method="class")
```
Note: to view the decision tree with fancy run this command:
```{r}
fancyRpartPlot(modFitA1)
```
  Predicting:
```{r}  
predictionsA1 <- predict(modFitA1, myTesting, type = "class")
```
Using confusion Matrix to test results:
```{r}
confusionMatrix(predictionsA1, myTesting$classe)
```

Using ML algorithms for prediction: Random Forests
```{r}
modFitB1 <- randomForest(classe ~. , data=myTraining)
```

Predicting in-sample error:
```{r}
predictionsB1 <- predict(modFitB1, myTesting, type = "class")
```
Using confusion Matrix to test results:
```{r}
confusionMatrix(predictionsB1, myTesting$classe)
```
Random Forests yielded better Results, as expected!


Generating Files to submit as answers for the Assignment:
Finally, using the provided Test Set out-of-sample error.

For Random Forests we use the following formula, which yielded a much better prediction in in-sample:
```{r}
predictionsB2 <- predict(modFitB1, testing, type = "class")
```
Function to generate files with predictions to submit for assignment
```{r}
pml_write_files = function(x){
  n = length(x)
  for(i in 1:n){
    filename = paste0("problem_id_",i,".txt")
    write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
  }
}

pml_write_files(predictionsB2)
```






